Investigating Fire Incidents Severity Levels in Toronto: Enhancing Resource Allocation for Fire Response

### 2024-12-04 paper/paper.qmd
paper regenrate

---
title: "Analyzing Toronto Fire Incident Severity"
subtitle: "Study Thorugh Financial and Casualty Loss to Optimize Resource Allocation for Toronto Fire Services"
author: 
  - Maggie Zhang
thanks: "Code and data are available at: [https://github.com/MaggieZ111119/TFS_Toronto_Fire_Incident](https://github.com/MaggieZ111119/TFS_Toronto_Fire_Incident)."
date: today
date-format: long
abstract: "This study investigates different severity levels of Toronto fire incidents to support efficient resource allocation for Toronto Fire Service (TFS). The analysis reveals that higher-severity incidents correlate with greater financial losses and more significant human casualties, impacting both civilians and firefighters. The study identifies high-priority response criteria, including incidents originating from XXXX regions, occurring within defined time frames: XXXX, and involving XXXX as the ignition sources. These insights, derived from the modeling analysis, can guide TFS to enhance response times and prioritize tasks based on common high-risk indicators. This work aims to inform decision-making and optimize fire response operations to improve community safety in Toronto."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

fire_data <- read.csv(here::here("~/TFS_Toronto_Fire_Incident/data/02-analysis_data/tfs_analysis_data"))

@@

This study investigates different severity levels of Toronto fire incidents based on financial and casualty losses to support efficient resource allocation for Toronto Fire Service (TFS). The analysis reveals that higher-severity incidents correlate with greater financial losses and more significant human casualties, impacting both civilians and firefighters.

The research identifies specific high-priority response criteria that can improve TFS operations. By examining the data, it has been found that incidents originating from XXX regions, occurring within defined time frames: XXX, and involving XXXX ignition sources exhibit higher risk profiles. This enables TFS to develop strategic response protocols that prioritize these types of incidents. The modeling analysis conducted in this study provides valuable insights into which factors contribute most significantly to incident severity, equipping TFS with the knowledge to enhance response times and allocate resources more effectively.

The insights from this work aim to inform strategic decision-making processes and optimize fire response operations. By understanding and addressing the indicators of high-risk incidents, TFS can better safeguard the community and improve public safety throughout Toronto. This research is an essential step toward developing data-driven approaches that enhance both the efficiency of fire service operations and the overall resilience of the city.

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

We use the statistical programming language R [@citeR]to investigate data about fire incidents[@TFSfireedata], downloaded from the Open Data Toronto Portal: https://open.toronto.ca/dataset/fire-incidents/. The dataset was last updated on December 31, 2023. It includes fire incidents following the Ontario Fire Marshal's (OFM) definition, with descriptions according to OFM standards. The dataset also provides detailed information on Toronto Fire Services' (TFS) responses to each fire incident. The data is mostly complete, though some incidents are still under investigation and categorized as "no loss outdoor fire." The dataset includes information on the origin of fires, impacts on civilians and buildings, as well as geographical data. Additionally, it contains details on TFS response times and actions taken. Overall, the dataset is comprehensive, informative, and provides valuable insights for my later study.

@@

This structured approach allows us to analyze and understand various aspects of fire incidents in Toronto. All variables in the dataset have the potential to provide valuable insights and perspectives for various analyses, contributing to fire prevention and safety improvements. However, for my analysis of fire incidents across Toronto, certain variables; such as latitude and longitude, incident numbers in the CAD system, Incident_Ward, and road intersections—are being excluded from the study. A more detailed explanation will be provided in the Appendix.

@@

To maintain data quality, rows with more than 30% missing data are removed. After ensuring that there are no missing values in the dataset, a new variable, Response_Time, is created to calculate the time difference (in minutes) between TFS_Alarm_Time and TFS_Arrival_Time. This new variable will help analyze fire response efficiency and is a key indicator for the study.
(@fig-area-origin-distribution).

```{r}
#| label: fig-area-origin-distribution
#| fig-cap: Distribution of the Area of Origin for fire incidents, categorized into various groups such as Residential, Office, Mechanical, and Industrial areas. The plot highlights the frequency of each area category, helping to understand the distribution of fire origins.
#| echo: false

all_area <- unique(fire_data$area_of_origin)
# List of Area of Origin
area_of_origin <- c("22 - Sleeping Area or Bedroom (inc. patients room, dormitory, etc)", 
#| echo: false

@@

  # Public and Performance Spaces
  else if (grepl("Stage|Rink|Gym|Performance|Concourse|Court", area)) {
    return("Public and Performance Spaces")
  } 
  # Miscellaneous and Unclassified
  else {
    return("Miscellaneous and Unclassified")
  }
}
# Apply the categorization function to each area of origin
fire_data$area_of_origin_grouped <- sapply(fire_data$area_of_origin, 
                                           categorize_area_of_origin)
# Plot
ggplot(fire_data, aes(x = area_of_origin_grouped)) +
  geom_bar(fill = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Area of Origin Distribution") +
  xlab("Area of Origin") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0.5)

@@

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

**Understanding Variables of Interest**:\
To understand the distribution of various categorical and numerical variables in the fire incident dataset, I visualized each variable along with its distribution and reviewed its summary statistics. The goal is to gain insights into the key characteristics of the data, which will support the later stages of analysis and modeling. Below are a few important variables considered:

1.  **Area of Origin**:\
    This is a categorical variable that indicates the location within a building or structure where a fire started. Initially, this variable had over 100 categories, so the goal was to categorize these locations into broader groups for easier interpretation and analysis. I grouped the areas into categories such as "Residential and Living Spaces," "Cooking and Dining Areas," and "Storage and Utility Areas," among others. These groups were then visualized to better understand the most common fire origins within the dataset (see Table 1: \[code\]).

2.  **Civilian Casualties**:\
    This is a numerical variable that provides insight into the severity of the incidents. Basic descriptive statistics (e.g., mean, median, and range) were calculated using the `summary()` function. Additionally, a scatter plot was created to explore the relationship between the number of civilian casualties and the incident ID (see \[code\]).

3.  **Final Incident Type**:\
    This is a categorical variable that classifies the type of incident (e.g., fire, explosion). A bar plot was created to visualize the distribution of different incident types. The frequency of each incident type is also displayed in the table (see \[code\]).

4.  **Ignition Source**:\
    This variable provides information about the cause of the fire. Categories of ignition sources, such as "Incandescent Lamp" or "Vehicle - Mechanical," were listed. The distribution of ignition sources was visualized with a bar plot (see \[code\]).

**Understanding Potential Indicators of Severity and Their Replashioships**:\
To explore the relationship between financial losses and casualties, incidents were grouped into financial loss and casualty categories. A heatmap below illustrates the interaction between these two indicators.

.

```{r}
fire_data <- fire_data %>%
  mutate(
    loss_group = cut(
      estimated_dollar_loss,
      breaks = c(0, 1000, 10000, 50000, 500000, Inf),
      labels = c("<1k", "1k-10k", "10k-50k", "50k-500k", ">500k"),
      right = FALSE
    ),
    casualty_group = cut(
      total_casualties,
      breaks = c(0, 1, 2, 5, Inf),
      labels = c("0", "1", "2-5", ">5"),
      right = FALSE
    )
  )

# Summarize data and plot heatmap
severity_table <- fire_data %>%
  group_by(loss_group, casualty_group) %>%
  summarize(count = n(), .groups = "drop")

ggplot(severity_table, aes(x = loss_group, y = casualty_group, fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Heatmap of Financial Loss and Casualty Groups",
    x = "Financial Loss Group",
    y = "Casualty Group",
    fill = "Incident Count"
  ) +
  theme_minimal()

@@

The scatter plot below showcases the relationship between the number of responding apparatus and the logarithmic scale of financial losses. A Pearson correlation coefficient was calculated to assess the strength of this relationship.

```{r}
fire_data$Log_Dollar_Loss <- log(fire_data$estimated_dollar_loss + 1)

# Scatter plot and correlation
ggplot(fire_data, aes(x = Log_Dollar_Loss, y = number_of_responding_apparatus)) +
  geom_point(color = 'blue', alpha = 0.5) +
  labs(
    title = 'Number of Responding Apparatus vs Log of Financial Loss',
    x = 'Log of Financial Loss',
    y = 'Number of Responding Apparatus'
  ) +
  theme_minimal()

# Calculate correlation
cor(fire_data$Log_Dollar_Loss, fire_data$number_of_responding_apparatus, use = "complete.obs")
```

# Model

@@

@@

The analysis shows that most incidents fall under the Low Severity category, which aligns with the reality of everyday fire service responses.

### Set Up Model For Predicting Severities

@@

To evaluate model performance, the dataset was split into a training set (80%) and a testing set (20%) using the `createDataPartition()` function from the `caret` package in R.

### Model Explnation

@@

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

>>>>

---
title: "Investigating Fire Incidents Severity Levels in Toronto: Enhancing Resource Allocation for Fire Response"
subtitle: "Identifying Key Factors for Fire Severity and Evaluating the Effectiveness of Public Fire Response Systems"
author: 
  - Maggie Zhang
thanks: "Code and data are available at: [https://github.com/MaggieZ111119/TFS_Toronto_Fire_Incident](https://github.com/MaggieZ111119/TFS_Toronto_Fire_Incident)."
date: "December 3, 2024"
date-format: long
abstract: "This study investigates fire severity levels in Toronto to improve resource allocation for Toronto Fire Services (TFS). The analysis identifies that higher-severity incidents are strongly correlated with incidents originating from mechanical, HVAC, and electrical areas, while cooking and lighting areas tend to lead to lower severity fires. The study also finds that fire incidents involving certain ignition sources, such as electrical units or lightning equipment, are more likely to be severe. The study also evaluates the effectiveness of public fire response systems, highlighting that fire alarms are successful in low and medium-severity fires, while sprinkler systems are often only partially operational. Results can guide TFS to enhance response times, prioritize resources, and optimize fire response operations, ultimately improving community safety in Toronto."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

fire_data <- read.csv(file = here::here("data/02-analysis_data/tfs_analysis_data"))

@@

The study aims to estimate the severity of fires in the Toronto region. A fire incident, as defined by the Ontario Fire Marshal (OFM), refers to any event involving fire, explosion, or other situations suspected to lead to significant outcomes, such as injury or death [@ontario_fire_marshal].

The severity of a fire can be assessed through various factors, including the intensity of the flames, the size of the affected area, the duration of the fire, its impact on human activities, and its potential threat to animal habitats. There are numerous standards and considerations when discussing fire severity, but when it comes to prediction, it is unrealistic to consider all factors. Therefore, defining severity becomes the first task before any further analysis. In this paper, the severity of a fire is defined based on three aspects: casualty and financial loss, the type of incident, and the number of apparatuses dispatched by Toronto Fire Services (TFS) in response to the fire. Further explanation of how severity is defined will be provided later in the paper.

This research also identifies specific high-priority response criteria that can enhance TFS operations. By examining the data, it has been found that the area of origin of a fire is a significant predictor of fire severity. Fires originating from mechanical, HVAC, and electrical areas are highly correlated with severe incidents, whereas fires originating from cooking and heating areas are more commonly associated with low-severity incidents.

Additionally, beyond investigating potential predictors of severe fire incidents, this study also explores the effectiveness of current public fire response mechanisms, such as fire alarms and sprinkler systems, in the Toronto region. It was discovered that sprinkler systems, across all levels of fire severity, were largely only partially operational. On the other hand, fire alarm systems were successfully activated in low and medium-severity fire incidents.

The Office of the Fire Marshal (OFM) plays a crucial role in overseeing fire safety regulations in Ontario, ensuring fire protection, prevention, and public safety across the province. They are responsible for setting safety standards and improving public safety for all municipalities. Under the leadership of the OFM, municipal fire departments, such as Toronto Fire Services (TFS), allocate resources and develop plans to protect shared communities [@ontario_fire_marshal]. Therefore, by analyzing data within their standards, identifying factors that contribute to high-risk fire incidents, and evaluating current operational mechanisms, TFS can gain valuable insights to develop strategic response protocols that prioritize these incidents. This knowledge will help both the public and fire departments enhance response times, allocate resources more effectively, and reduce the occurrence of lower-severity incidents.

The remainder of this paper is structured as follows. [@sec-data] describes the data used in the analysis, including the sources and key variables and their visualizations.[@sec-model] outlines the models applied to estimate fire severity and identify key factors. [@sec-results] presents the results of the analysis. [@sec-discussion] provides an interpretation of the results in the context of current fire safety practices and suggests potential improvements. Finally, [@sec-appendixa] includes supplementary information and additional data details. [@sec-appendix-B] provided insight into Fire Incidence, about their sampling method and potential bias.

In this paper, I used several R packages for data analysis and visualization. The `tidyverse`[@wickham2023tidyverse], `dplyr`[@wickham2023dplyr], and `tidyr`[@wickham2023tidyr] were used for data manipulation. `ggplot2`[@wickham2016ggplot2] facilitated data visualization, while `lubridate`[@grolemund2011lubridate] was used for date-time handling. `knitr`[@xie2023knitr] helped integrate code into reports. For modeling, I used `rstanarm`[@gelman2020rstanarm] and `modelsummary`[@fox2021modelsummary] for summaries, `caret`[@kuhn2023caret] for model training, and `xgboost`[@chen2016xgboost] for boosting. Additional tools like `rpart`[@therneau2015rpart], `fitdistrplus`[@delignette-muller2023fitdistrplus], and `pROC`[@robin2023pROC] supported partitioning, distribution fitting, and model evaluation.

@@

We use the statistical programming language R [@citeR]to investigate data about fire incidents[@TFSfireedata], downloaded from the Open Data Toronto Portal: https://open.toronto.ca/dataset/fire-incidents/ . The dataset was last updated on December 31, 2023. It includes fire incidents following the Ontario Fire Marshal's (OFM) definition, with descriptions according to OFM standards. The dataset also provides detailed information on Toronto Fire Services' (TFS) responses to each fire incident. The data is mostly complete, though some incidents are still under investigation and categorized as "no loss outdoor fire." The dataset includes information on the origin of fires, impacts on civilians and buildings, as well as geographical data (see @tbl-preview-data). Additionally, it contains details on TFS response times and actions taken. Overall, the dataset is comprehensive, informative, and provides valuable insights for my later study.

```{r}
#| label: tbl-preview-data
#| tbl-cap: Preview of the Fire Incident data (https://open.toronto.ca/dataset/fire-incidents/) before cleaning, displaying the first six rows with columns decided first in interested for study.
#| echo: false
#| warning: false

library(tidyverse)
library(dplyr)
library(lubridate)

#### Load original data ####
raw_data <- read_csv(file = here::here("data/01-raw_data/raw_fire_data.csv"))

columns_to_keep <- c("_id", "Area_of_Origin", "Building_Status", "Business_Impact", 
                     "Civilian_Casualties", "Estimated_Dollar_Loss", "Exposures", 
                     "Final_Incident_Type", "Fire_Alarm_System_Operation", 
                     "Ignition_Source",
                     "Number_of_responding_apparatus", "Number_of_responding_personnel", 
                     "Possible_Cause", "Sprinkler_System_Presence", 
                     "TFS_Alarm_Time", "TFS_Arrival_Time", "TFS_Firefighter_Casualties")

clean_data <- raw_data[, columns_to_keep]
# Change columns name to lower case and ensure no space
colnames(clean_data) <- tolower(gsub(" ", "_", colnames(clean_data)))

### Preview data ###
kable(head(clean_data, 6))
```

@@

This structured approach allows us to analyze and understand various aspects of fire incidents in Toronto. All variables in the dataset have the potential to provide valuable insights and perspectives for various analyses, contributing to fire prevention and safety improvements. However, for my analysis of fire incidents across Toronto, certain variables; such as latitude and longitude, incident numbers in the CAD system, Incident_Ward, and road intersections—are being excluded from the study. A more detailed explanation will be provided in the @sec-appendixa.

@@

To maintain data quality, rows with more than 30% missing data are removed. After ensuring that there are no missing values in the dataset, a new variable, Response_Time, is created to calculate the time difference (in minutes) between TFS_Alarm_Time and TFS_Arrival_Time (see @fig-response-time-distribution). This new variable will help analyze fire response efficiency and is a key indicator for the study.

```{r}
#| label: fig-response-time-distribution
#| fig-cap: Distribution of response times for fire incidents in Toronto, showing the frequency of various response times (in minutes). This visualization highlights the efficiency of the Toronto Fire Service's (TFS) response. It's almost normally distributed with a mean of around 5, and the spread of not too far.
#| echo: false
#| warning: false
library(ggplot2)
fire_data <- read.csv(here::here("data/02-analysis_data/tfs_analysis_data"))
## 15. Response Time (Numerical) ##
#Plot
ggplot(fire_data, aes(x = response_time)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  theme_minimal() +
  ggtitle("Response Time Distribution") +
  xlab("Response Time (minutes)") +
  ylab("Frequency") +
  stat_bin(aes(label = ..count..), geom = "text", vjust = 0.5, size = 3) +
  xlim(0, 15)

**Understanding Variables of Interest**:\
To understand the distribution of various categorical and numerical variables in the fire incident dataset, I visualized each variable along with its distribution and reviewed its summary statistics. The goal is to gain insights into the key characteristics of the data, which will support the later stages of analysis and modeling. Can check @sec-appendixa for visualization/summary of the distribution of all variables. Below are a few important variables considered:

1.  **Area of Origin**:\
    This is a categorical variable that indicates the location within a building or structure where a fire started. Initially, this variable had over 100 categories, so the goal was to categorize these locations into broader groups for easier interpretation and analysis. I grouped the areas into categories such as "Residential and Living Spaces," "Cooking and Dining Areas," and "Storage and Utility Areas," among others. These groups were then visualized to better understand the most common fire origins within the dataset (see @fig-area-origin-distribution). Among all categories, Cooking and Dinning area stands out to be the area which most fire originate from. Living space and Transportation area also has a great amount to incidence occured.

```{r}
#| label: fig-area-origin-distribution
#| fig-cap: Distribution of the Area of Origin for fire incidents, categorized into various groups such as Residential, Office, Mechanical, and Industrial areas. The plot highlights the frequency of each area category, helping to understand the distribution of fire origins.
#| echo: false
#| warning: false

@@

  # Public and Performance Spaces
  else if (grepl("Stage|Rink|Gym|Performance|Concourse|Court", area)) {
    return("Public and Performance Spaces")
  } 
  # Miscellaneous and Unclassified
  else {
    return("Miscellaneous and Unclassified")
  }
}
# Apply the categorization function to each area of origin
fire_data$area_of_origin_grouped <- sapply(fire_data$area_of_origin, 
                                           categorize_area_of_origin)
# Plot
ggplot(fire_data, aes(x = area_of_origin_grouped)) +
  geom_bar(fill = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Area of Origin") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0.5)

@@

2.  **Civilian Casualties**:\
    This is a numerical variable that provides insight into the severity of the incidents. Basic descriptive statistics (e.g., mean, median, and range) were calculated using the `summary()` function. Additionally, a scatter plot was created to explore the relationship between the number of civilian casualties and the incident ID (see @tbl-civilian-casualties-summary). We can see that most of the time, there are no casualties. This data set ranges across over 10 years (see @fig-fire-incidents-by-year in @sec-appendixa), looking at all levels of casualties, and taking its average over 6 years, we can see that this indicates the city has done a good job in protecting civilians from fire places.

```{r}
#| label: tbl-civilian-casualties-summary
#| tbl-cap: Summary table showing the number of civilian casualties in fire incidents, including the frequency of each casualty count. This table summarizes the distribution of civilian casualties and provides an overview of the impact of fire incidents on civilians.
#| echo: false
#| warning: false
library(knitr)

#Overall Table
civilian_casualties_df <- as.data.frame(table(fire_data$civilian_casualties))
colnames(civilian_casualties_df) <- c("Number_of_Casualties", "Frequency")
kable(civilian_casualties_df)

@@

3.  **Final Incident Type**:\
    This is a categorical variable that classifies the type of incident (e.g., fire, explosion). A bar plot was created to visualize the distribution of different incident types.(see @fig-final-incident-type-distribution). It's clear that there are more fire than explotion.

```{r}
#| label: fig-final-incident-type-distribution
#| fig-cap: Bar plot illustrating the distribution of final incident types across fire incidents. The plot presents the frequency of each incident type, with the count displayed on top of the bars. We can see that there is much more fire than explotion.
#| echo: false

ggplot(fire_data, aes(x = final_incident_type)) +
  geom_bar(fill = "purple") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    plot.margin = margin(10, 10, 10, 10)  # Add space around the plot
  ) +
  scale_x_discrete(labels = function(x) sapply(x, function(y) paste(head(strsplit(y, " ")[[1]], 3), collapse = " "))) +
  xlab("Final Incident Type") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0.5)

4.  **Ignition Source**:\
    This variable provides information about the cause of the fire. Categories of ignition sources, such as "Incandescent Lamp" or "Vehicle - Mechanical," were listed. The distribution of ignition sources was visualized with a bar plot (see @fig-grouped-ignition-source-distribution). We can tell that fire-related and open flame is one of the biggest source of fire incidence; cooking and heating equipment are second place as sources, which align with @fig-area-origin-distribution result, where Ketchen is the place most fire incidence starts off.
#| label: fig-grouped-ignition-source-distribution
#| fig-cap: Bar plot illustrating the distribution of grouped ignition sources for fire incidents. The plot categorizes ignition sources into meaningful groups such as Lighting and Electrical Equipment, Cooking and Heating Equipment, and Mechanical Equipment, among others. The frequency of each group is displayed. Fire-related and Open Flame is the bigget ingnition source.
#| echo: false
#| warning: false
all_sources <- unique(fire_data$ignition_source)
ignition_sources <- c("51 - Incandescent Lamp - Light Bulb, Spotlight", 
                      "23 - Distribution Equipment (includes panel boards, fuses, circuit br)", 
                      "41 - Other Heating Equipment", "11 - Stove, Range-top burner", 
                      "82 - Vehicle - Mechanical", "24 - Circuit Wiring - Copper", 
                      "98 - Other", "81 - Vehicle - Electrical", "49 - Other Appliances", 
                      "71 - Smoker's Articles (eg. cigarettes, cigars, pipes already ignited)", 
                      "13 - Microwave", "77 - Matches or Lighters (unable to distinguish)", 
                      "74 - Salamander", "12 - Oven", "47 - Refrigerator, Freezer (includes vending machine)", 
                      "73 - Blow Torch, Bunsen Burner", "72 - Cutting/Welding Equipment", 
                      "21 - Transformer", "55 - Candle", "29 - Extension Cord, Temporary Wiring", 
                      "14 - Open Fired Barbeque - Fixed or Portable", "79 - Other Open Flame Tools/Smokers' Articles", 
                      "84 - Other Mechanical", "36 - Fireplace - Masonry", "83 - Other Electrical", 
                      "20 - Service/Utility Lines (includes power/hydro transmission lines)", 
                      "92 - Open Fire (eg. camp fire, rubbish fire, etc.)", "17 - Wood burning stove", 
                      "62 - Heat Treatment Equipment (eg. furnace, oven, kiln, quench tanks, etc.)", 
                      "19 - Other Cooking Items (eg Toaster, Kettle, elec frying pan)", 
                      "93 - Hot Ashes, Embers, Spark", "52 - Florescent Lamp (includes ballast)", 
                      "64 - Chemical Processing Equipment (eg. reactors, distilling units, etc.)", 
                      "43 - Clothes Dryer", "69 - Other Processing Equipment", 
                      "88 - Multiple Ignition Source or Igniting Equipment (suspected arson)", 
                      "75 - Matches (open flame)", "32 - Water Heater", 
                      "26 - Terminations-Copper (incl receptacles, switches, lights)", "85 - Vehicle collision", 
                      "28 - Cord, Cable for Appliance, Electrical Articles", "16 - Deep Fat Fryer", 
                      "59 - Other Lighting Equipment", "30 - Other Electrical Distribution Item", 
                      "76 - Lighters (open flame)", "33 - Space Heater - Fixed", "42 - Television, Radio, Stereo, Tape Recorder, etc.", 
                      "39 - Chimney - Masonry", "34 - Space Heater - Portable", "44 - Iron, Pressing Machine", 
                      "108 - Exposure, source other", "15 - Range Hood", "63 - Painting Equipment", 
                      "107 - Exposure, source vehicle (outside structure)", 
                      "104 - Exposure, source open fire (inc campfire, rubbish fire)", 
                      "48 - Air Conditioner - Window or Room Unit", "96 - Chemical Reaction (eg. spontaneous combustion, etc.)", 
                      "106 - Exposure, source grass, shrubs, trees", "35 - Fireplace - Factory Built", 
                      "37 - Fireplace Insert", "94 - Static Electricity (spark)", "9990 - Under Investigation", 
                      "31 - Central Heating/Cooling Unit", "56 - Halogen Lamp or light", 
                      "91 - Fireworks", "46 - Electric Blanket, Heating Pad", 
                      "105 - Exposure, source forest, trees, wildland", "100 - Outdoor fireplace/heater", 
                      "103 - Exposure, source outside storage container, tank", 
                      "38 - Chimney - Factory Built", "54 - Lamp (eg. coal, oil, naphtha, etc.)", 
                      "102 - Exposure, source structure semi-detached or attached", "40 - Flue Pipe", 
                      "27 - Terminations-Aluminum (incl receptables, switches, lights)", 
                      "22 - Meter", "80 - Portable generator", "101 - Exposure, source structure detached", 
                      "95 - Lightning", "97 - Rekindle", "45 - Washing Machine", 
                      "25 - Circuit Wiring - Aluminum", "53 - Christmas Lights, Decorative Lighting", 
                      "61 - Incinerator", "90 - Explosives")

# Create a function to categorize sources
categorize_ignition_sources <- function(source) {
  # Lighting and Electrical Equipment
  if (grepl("Lamp|Lighting|Electrical|Cable|Wire|Ballast|Transformer", source)) {
    return("Lighting and Electrical Equipment")
  } 
  # Cooking and Heating Equipment
  else if (grepl("Stove|Oven|Microwave|Grill|Fryer|Heater|Range", source)) {
    return("Cooking and Heating Equipment")
  } 
  # Fire-related and Open Flame
  else if (grepl("Fire|Matches|Candle|Flame|Torch|Burner|Smoker", source)) {
    return("Fire-related and Open Flame")
  } 
  # Flammable Materials and Chemicals
  else if (grepl("Chemical|Explosion|Flammable|Rekindle|Static|Spark", source)) {
    return("Flammable Materials and Chemicals")
  } 
  # Mechanical Equipment
  else if (grepl("Mechanical|Vehicle|Collision|Smokers|Blower", source)) {
    return("Mechanical Equipment")
  } 
  # Electrical Distribution and Utility
  else if (grepl("Distribution|Service|Utility|Panel|Circuit", source)) {
    return("Electrical Distribution and Utility")
  }

  # Static Electricity and Sparks
  else if (grepl("Spark|Static", source)) {
    return("Static Electricity and Sparks")
  } 
  # Exposure to Fire and Environmental
  else if (grepl("Exposure|Outside|Container|Fireplace|Campfire|Wildland", source)) {
    return("Exposure to Fire and Environmental")
  } 
  # Processing Equipment (Industrial)
  else if (grepl("Furnace|Kiln|Processing|Incinerator", source)) {
    return("Processing Equipment")
  } 
  # Appliances
  else if (grepl("Appliance|Refrigerator|Washing Machine|Dryer|Iron", source)) {
    return("Appliances")
  }

  # Vehicle and Accidents
  else if (grepl("Vehicle|Accident|Collision", source)) {
    return("Vehicle and Accidents")
  } 
  # Multiple Ignition Sources (or suspected arson)
  else if (grepl("Multiple|Arson", source)) {
    return("Multiple Ignition Sources")
  } 
  # Cooling and Air Conditioning Equipment
  else if (grepl("Cooling|Air Conditioner", source)) {
    return("Cooling and Air Conditioning Equipment")
  } 
  # Other (for anything that doesn't fit into the above categories)
  else {
    return("Other")
  }
}
# Apply the categorization function to each ignition source
fire_data$ignition_source_grouped <- sapply(fire_data$ignition_source, categorize_ignition_sources)
#Table
#Plot
# Plot of grouped ignition source distribution
ggplot(fire_data, aes(x = ignition_source_grouped)) +
  geom_bar(fill = "green") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("Ignition Source Group") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0.5)

**Understanding Potential Indicators of Severity and Their Replashioships**:\
To explore the relationship between financial losses and casualties, incidents were grouped into financial loss and casualty categories. A heatmap below (@fig-financial-loss-vs-casualties-heatmap) illustrates the interaction between these two indicators.

```{r}
#| label: fig-financial-loss-vs-casualties-heatmap
#| fig-cap: Heatmap illustrating the relationship between financial loss and casualty groups in fire incidents. The plot categorizes incidents based on financial loss and total casualties, with the color intensity representing the frequency of incidents in each combination of loss and casualty group. This provides insights into the severity of fire incidents, highlighting areas with higher financial losses and casualties. We can see that there are greater anount in 1k -10k region.
#| echo: false
#| warning: false

## Financial loss vs. Casualties ##
fire_data$total_casualties <- fire_data$civilian_casualties + fire_data$tfs_firefighter_casualties

#heatmap
fire_data <- fire_data %>%
  mutate(
    loss_group = cut(
      estimated_dollar_loss,
      breaks = c(0, 1000, 10000, 50000, 500000, Inf),
      labels = c("<1k", "1k-10k", "10k-50k", "50k-500k", ">500k"),
      right = FALSE
    ),
    casualty_group = cut(
      total_casualties,
      breaks = c(0, 1, 2, 5, Inf),
      labels = c("0", "1", "2-5", ">5"),
      right = FALSE
    )
  )

severity_table <- fire_data %>%
  group_by(loss_group, casualty_group) %>%
  summarize(count = n())

ggplot(severity_table, aes(x = loss_group, y = casualty_group, fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    x = "Financial Loss Group",
    y = "Casualty Group",
    fill = "Incident Count"
  ) +
  theme_minimal()

The scatter plot below showcases the relationship between the number of responding apparatuses and the logarithmic scale of financial losses (@fig-responding-apparatus-vs-log-dollar-loss). We can observe a slight trend between responding apparatus and financial loss A Pearson correlation coefficient was calculated to assess the strength of this relationship.

```{r}
#| label: fig-responding-apparatus-vs-log-dollar-loss
#| fig-cap: Scatter plot showing the relationship between the number of responding apparatus and the log-transformed financial loss in fire incidents. Each point represents a fire incident, with the x-axis displaying the log of the financial loss (with a 1 added for stability) and the y-axis showing the number of responding apparatus. There is a slight trend observed, indicating potential relationship.
#| echo: false

fire_data$Log_Dollar_Loss <- log(fire_data$estimated_dollar_loss + 1)

# Scatter plot between number of responding apparatus and final dollar loss
ggplot(fire_data, aes(y = number_of_responding_apparatus, x = Log_Dollar_Loss)) +
  geom_point(color = 'blue', alpha = 0.5) +
  labs(
       x = 'Number of Responding Apparatus',
       y = 'Log of Final Dollar Loss') +
  theme_minimal()

# cor(fire_data$Log_Dollar_Loss, fire_data$number_of_responding_apparatus, use = "complete.obs")

# Model {#sec-model}

@@

fire_data_model <- read_csv( file = here::here("data/02-analysis_data/severity_tfs_analysis_data"))

@@

The analysis shows that most incidents fall under the Low Severity category (see @tbl-severity-count), which aligns with the reality of everyday fire service responses.

```{r}
#| label: tbl-severity-count
#| tbl-cap: The table shows the distribution of fire incidents across Low, Medium, and High severity levels based on the number of responding apparatus, estimated dollar loss, and total casualties and incident type. Notice that there are musch more lower severe incidence.
#| echo: false
#| warning: false
#### Workspace setup ####
library(tidyverse)
library(rstanarm)
library(rpart)
library(ggplot2)
library(dplyr)
library(tidyr)
library(fitdistrplus)
library(knitr)

#### Read data ####
fire_data_model <- read_csv(file = here::here("data/02-analysis_data/updated_tfs_analysis_data"))

#### Quantify Severity ####

### Apparatus
fire_data_model <- fire_data_model %>%
  mutate(severity_bin = cut(number_of_responding_apparatus,
                            breaks = c(0, 10, 20, 30, 40, 50, Inf),
                            labels = c("[1,10]", "[11,20]", "[21,30]", "[31,40]", "[41,50]", ">50"),
                            right = TRUE))
severity_dist <- fire_data_model %>%
  group_by(severity_bin) %>%
  summarise(frequency = n()) %>%
  ungroup()
#severity_dist

### Estimated_dollar_loss
# Log transform: 
fire_data_model$log_estimated_dollar_loss <- log(fire_data_model$estimated_dollar_loss + 1)
# Recalculate quantiles on transformed data
loss_quantiles <- quantile(fire_data_model$log_estimated_dollar_loss, probs = c(0.25, 0.5, 0.75))

### Determining Severity
fire_data_model <- fire_data_model %>%
  mutate(
    # Define Casualty Severity
    casualty_severity = case_when(
      total_casualties == 0 ~ "Low", 
      total_casualties >= 4 ~ "High",
      TRUE ~ "Medium"
    ),

    # Define Final Severity based on Apparatus, Dollar Loss, and Casualties
    Severity = case_when(
      # Explosion = High severity
      final_incident_type == 
        "02 - Explosion (including during Fire, excluding Codes 3 & 11-13)"
      ~ "High",

      # Apparatus severity classifications based on bins
      (casualty_severity == "Low" 
       | severity_bin %in% c("[1,10]") 
       | log_estimated_dollar_loss <= loss_quantiles[1]) ~ "Low",

      (casualty_severity == "Medium" 
       | severity_bin %in% c("[11,20]", "[21,30]") 
       | log_estimated_dollar_loss <= loss_quantiles[2]) ~ "Medium",

      (casualty_severity == "High" 
       | severity_bin %in% c("[31,40]", "[41,50]", ">50") 
       | log_estimated_dollar_loss > loss_quantiles[2]) ~ "High",

      TRUE ~ "Low" 
    )
  )
severity_count_table <- fire_data_model %>%
  count(Severity)

# Display the table
kable(severity_count_table)
```

## Set Up Model For Predicting Severities

@@

To evaluate model performance, the dataset was split into a training set (80%) and a testing set (20%) using the `createDataPartition()` function from the `caret` package in R [@citeR].

## Model Explnation

@@

logistic_model <-
  readRDS(file = here::here("models/logtsic_model.rds"))

library(ggplot2)
coef_df <- data.frame(
  Area_of_Origin = names(coef(logistic_model))[-1],
  Coefficients = coef(logistic_model)[-1]
)
